Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-08-08 paper/main.tex
Updates

\begin{enumerate}
	\item \textbf{Continuous:} We train a regression
		model, $ E_X: x \sim \bm{z} $. The residuals are then defined
		as the difference between the true and the predicted value using
		$ E_X $.
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability
		estimator, $ p_X: x \sim \bm{z} $, and take probability
		predictions from it $ \hat{p}_X: \hat{p}_X(\bm{Z}) $. Then we
		compute the residuals as follows:
		$$ R_{x_i} = \hat{p}(\hat{x}_i > x_i) - \hat{p}(\hat{x}_i < x_i) $$
	\item \textbf{Categorical:} We again start by training a
		probability estimator $ p_X: X \sim \bm{Z} $, and take the
		probability estimates from it $ \hat{p}_X: p_Z(\bm{Z}) $. Next,
		we one-hot encode the categorical variable and then compute the
		residuals as follows: 
		$$ R_X = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

@@

We repeat the above step for the variable $ Y $ and compute its residual $ R_Y
$. Here, we have the option to choose the estimators based on the properties of
their data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can work on any combination of data types while performing reasonably
well. For linear datasets we can also choose variants of linear regression
depending on the type of data. After the above steps are repeated for both $ X
$ and $ Y $, we end up with two residual matrices: $ R_X $ and $ R_Y $. The
type of the variable determines the shape for these matrices, i.e., if the
variable is continuous we get a residual matrix of shape $ ( n \times 1 ) $ and
if the variable is categorical we have a residual matrix of shape $ ( n \times
(k-1)) $ (where one of the dummy encoded variables is dropped to avoid
multicollinearity)

We now have two continuous valued residual matrices and the second step now is
to quantify the association between these residuals. In multivariate
statistics, canonical correlation based measures have been used for measuring
association between two sets of variables. We can treat the residuals as sets
of variables in case of categorical variable and use canonical correlations to
quantify the association between them.

@@

	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\phi(\bm{U}, \bm{V}) $ is defined as:

@@

\begin{equation}
	\begin{split}
		\bm{\rho}(\bm{U}, \bm{V}) &= \sqrt{\mathrm{E} \left( \Sigma_{\bm{UU}}^{-1} \Sigma_{\bm{UV}} \Sigma_{\bm{VV}}^{-1} \Sigma_{\bm{VU}} \right)}
	\end{split}
\end{equation}

\end{definition}

where $ \mathrm{E}(\Sigma) $ denotes the eigenvalues of the matrix $ \Sigma $.
$ \Sigma_{UU} $ and $ \Sigma_{VV} $ are the covariance matrices of $ \bm{U} $
and $ \bm{V} $, and $ \Sigma_{UV} $ and $ \Sigma_{VU} $ are the
cross-covariance matrices.

Canonical correlations are generalization of correlations to multi-dimensional
variables. It tries to find orthogonal linear transformations $ a $ and $ b $
of $ \bm{U} $ and $ \bm{V} $ such that the correlation between the transformed
variables $ a^T \bm{U} $ and $ b^T \bm{V} $ is maximized. It gives a vector of
correlation values of size $ \min(\rvert U \rvert, \rvert V \rvert) $
representing the correlation coefficient of each column of the transformed
variables. Pearson's correlation coefficient is a special case of canonical
correlations when $ \rvert U \rvert = \rvert V \rvert = 1$.

Many measures of association based on canonical correlation have been used in
multivariate statistics with each of them having slightly different properties.
\begin{itemize}
	\item Wilks' Lambda:
		$$ \Lambda = \prod (1 - \phi_i^2) $$
	\item Roy's Largest Root: 
		$$ \theta = \phi_{\max}^2 $$
	\item Pillai's Trace: 
		$$ \tau = \sum \phi_i^2 $$
\end{itemize}

We use Pillai's Trace for our purpose for two main reasons: 1) It uses all the
canoncial correlation values, 2) Its interpretation is similar to correlation
coefficient, i.e., $ 0 $ signifies no association and $ 1 $ signifies perfect
linear relationship. This gives us the following partial measure of association:

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ representing
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of sample size: }
	\item \textbf{Invariant to the dummy encoding scheme: } As we use dummy
		encoding to compute the residuals for categorical variables,
		but as canonical correlations search for linear combinations
		over the columns, the measure is invariant to the dummy
		encoding scheme that we use.
	\item \textbf{Generalizes partial correlations to mixed data: }
		As canonical correlation is equivalent to Pearson's correlation
		coefficient, when both $ X $ and $ Y $ are continuous, $ \rho $
		is equivalent to canonical correlations. Hence, $ \rho $ is a
		generalization of partial correlations to mixed data.
	\item \textbf{Related to Cramer\'s V: }
\end{enumerate}
\todo[inline]{Expand on these points}

>>>>

\begin{enumerate}
	\item \textbf{Continuous:} We train a regression model, $ E_X: x \sim
		\bm{z} $. The residuals are then defined as the difference
		between the true and the predicted values using $ E_X $. 
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability estimator, $
		p_X: x \sim \bm{z} $, and obtain probability predictions 
		$ \hat{p}_X: \hat{p}_X(\bm{Z}) $. Then, we compute the residuals
		as follows \citep{Li2012}:
		$$ R_{x_i} = \hat{p}(\hat{x}_i > x_i) - \hat{p}(\hat{x}_i < x_i) $$
	\item \textbf{Categorical:} We again start by training a probability
		estimator $ p_X: X \sim \bm{Z} $, and obtain probability
		estimates $ \hat{p}_X: p_Z(\bm{Z}) $. Next, we dummy encode the
		categorical variable and then compute the residuals as follows: 
		$$ R_X = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

@@

We repeat the above step for the variable $ Y $ and compute its residual $ R_Y
$. Here, we have the option to choose the estimators based on the properties of
the data. Non-parametric ensemble estimators such as Random Forest and XGBoost
can handle any combination of data types while performing reasonably well. For
linear datasets, we can also choose variants of linear regression depending on
the type of data.

After the above steps are repeated for both $ X $ and $ Y $, we end up with two
residual matrices: $ R_X $ and $ R_Y $. The type of the variable determines the
shape for these matrices. If the variable is continuous, we get a residual
matrix of shape $ ( n \times 1 ) $ and if the variable is categorical, we have
a residual matrix of shape $ ( x \times (k -1)) $ where one of the dummy encoded
variables is dropped to avoid multicollinearity.

We now have two continuous-valued residual matrices. The next step is
to quantify the association between these residuals. In multivariate
statistics, canonical correlation based measures have been used to measure
the association between two sets of variables. We can treat our residuals
as sets of variables and use canonical correlations to quantify the association
between them.

@@

	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\rho(\bm{U}, \bm{V}) $ is defined as:

@@

% \begin{equation}
% 	\begin{split}
% 		\bm{\rho}(\bm{U}, \bm{V}) &= \sqrt{\mathrm{E} \left( \Sigma_{\bm{UU}}^{-1} \Sigma_{\bm{UV}} \Sigma_{\bm{VV}}^{-1} \Sigma_{\bm{VU}} \right)}
% 	\end{split}
% \end{equation}
	\begin{equation}
		\rho(\bm{U}, \bm{V}) = \max_{a, b} Corr(a^T \bm{U}, b^T \bm{V})
	\end{equation}

	where $ a $ and $ b $ are vectors of coefficients that maximize the correlations
	between the linear combinations of $ a^T \bm{U} $ and $ b^T \bm{V} $.
\end{definition}

% where $ \mathrm{E}(\Sigma) $ denotes the eigenvalues of the matrix $ \Sigma $.
% $ \Sigma_{UU} $ and $ \Sigma_{VV} $ are the covariance matrices of $ \bm{U} $
% and $ \bm{V} $, and $ \Sigma_{UV} $ and $ \Sigma_{VU} $ are the
% cross-covariance matrices.

Canonical correlations generalize correlations to multi-dimensional variables. They
find orthogonal linear transformations $ a $ and $ b $ of $ \bm{U} $ and $ \bm{V} $
such that the correlation between the transformed variables $ a^T \bm{U} $ and 
$ b^T \bm{V} $ is maximized. This yields a vector of correlation values of 
size $ \min(p, q) $, representing the correlation coefficient of each pair of 
transformed variables. Pearson's correlation coefficient is a special case of 
canonical correlations when $ p = q = 1 $.

Many measures of association based on canonical correlation have been used in
multivariate statistics, each with slightly different properties.
\begin{itemize}
	\item Wilks' Lambda:
		$$ \Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $$
	\item Roy's Largest Root: 
		$$ \theta = \max_i(\phi_i^2) $$
	\item Pillai's Trace: 
		$$ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $$
\end{itemize}

We use Pillai's Trace for our purpose for two main reasons: 1) It uses all the
canonical correlation values, 2) Its interpretation is similar to correlation
coefficient, i.e., $ 0 $ signifies no association and $ 1 $ signifies perfect
linear relationship. This gives us the following partial measure of
association:

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ represents
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding Scheme: } Since canonical
		correlations search for linear combinations over the columns,
		the measure is invariant to the specific dummy encoding
		scheme used for categorical variables.
	\item \textbf{Generalizes Partial Correlations to Mixed Data: }
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Related to Cramer\'s V: } 
\end{enumerate}
